---
title: "How Member and Casual riders Bike Usage Differ"
author: "Haffizzan Ibrahim"
date: "1/19/2022"
output:
  pdf_document: default
  html_document: default
---
##set up environment

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readxl)
library(tinytex)

getwd()
setwd("D:/data_analytics/capstone_project/cs1/2021-divvy-tripdata-xsl")
```
## import data

```{r}
jan_21 <- read_xlsx("202101-divvy-tripdata.xlsx")
feb_21 <- read_xlsx("202102-divvy-tripdata.xlsx")
mar_21 <- read_xlsx("202103-divvy-tripdata.xlsx")
apr_21 <- read_xlsx("202104-divvy-tripdata.xlsx")
may_21 <- read_xlsx("202105-divvy-tripdata.xlsx")
jun_21 <- read_xlsx("202106-divvy-tripdata.xlsx")
jul_21 <- read_xlsx("202107-divvy-tripdata.xlsx")
aug_21 <- read_xlsx("202108-divvy-tripdata.xlsx")
sep_21 <- read_xlsx("202109-divvy-tripdata.xlsx")
oct_21 <- read_xlsx("202110-divvy-tripdata.xlsx")
nov_21 <- read_xlsx("202111-divvy-tripdata.xlsx")
dec_21 <- read_xlsx("202112-divvy-tripdata.xlsx")
```
## combine data into one tables
Data were made sure to have the same amount and name of variables and same character type for each variables, before combining them into one table.

Notes: colnames(), str(), and glimpse() functions were used to ensure the consistency of data.

```{r}
trips_21 <- bind_rows(jan_21, feb_21, mar_21, apr_21, may_21, jun_21,
                     jul_21, aug_21, sep_21, oct_21, nov_21, dec_21)

trips_21
```
## add new columns (date, day, month, year, day_of_week)

```{r}
trips_21$date <- format(as.Date(trips_21$start_time))
trips_21$year <- format(as.Date(trips_21$date), "%y")
trips_21$month <- format(as.Date(trips_21$date), "%m")
trips_21$day <- format(as.Date(trips_21$date), "%d")
trips_21$day_of_week <- format(as.Date(trips_21$date), "%A")
```
## find ride_length using difftime() function and add new colum

```{r}
trips_21$ride_length <- difftime(trips_21$end_time, trips_21$start_time)
trips_21$ride_length <- as.numeric(as.character(trips_21$ride_length))
```
## filter out rows with ride_length less than zero

```{r}
trips_21 <- trips_21 %>% filter(!ride_length < 0)

trips_21
```
# Descriptive analysis

```{r}
mean(trips_21$ride_length)
median(trips_21$ride_length)
max(trips_21$ride_length)
min(trips_21$ride_length)
summary(trips_21$ride_length)
```
## aggregate mean of ride length by user type and day of week

```{r}
aggregate(trips_21$ride_length ~
            trips_21$user_type +
            trips_21$day_of_week,
          FUN = mean)
```
## arrange day of week in order

```{r}
trips_21$day_of_week <- ordered(trips_21$day_of_week,
                                    levels=c("Sunday", "Monday",
                                             "Tuesday", "Wednesday",
                                             "Thursday", "Friday",
                                             "Saturday"))
```
## number of ride and average duration were compared based on user type

```{r}
outline_21 <- trips_21 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(user_type, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length), .groups = "drop") %>% 
  arrange(user_type, weekday)

outline_21
```
##visualization of findings

### Number of ride

```{r}
outline_21 %>% ggplot(aes(x = weekday, y = number_of_rides, fill = user_type)) +
  geom_col(position = "dodge")
```
### Average duration

```{r}
outline_21 %>% ggplot(aes(x = weekday, y = average_duration, fill = user_type)) +
  geom_col(position = "dodge")
```
